Making Sequential Programs Deterministic

### 2019-04-06 paper.tex

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
When a program is nondeterministic, it is difficult to test and debug.
Nondeterminism can occur even in sequential programs, for example as a
result of iterating over the elements of a hash table.

We have created a type system for determinism.  A type checker warns if a
program is nondeterministic or verifies that the program is deterministic.
The key ideas in the type system are type qualifiers for nondeterminism,
order-nondeterminism, and determinism; type well-formedness rules to
restrict the typings for collections; and enhancements to polymorphism that
improve precision when analyzing collection operations.

We implemented our type system for Java.  In a case study, it found
previously-unknown nondeterminism errors in a program that had been heavily
vetted by its developers.  In an experiment, it found all of the
non-concurrency-related nondeterminism that was found by other tools.

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
When a program is nondeterministic, it is difficult to test and debug.
Nondeterminism occurs even in sequential programs: for example, as a
result of iterating over the elements of a hash table. This seemingly innocuous and
frequently used operation can result in diverging test results
if the test depends on iteration order for asserting correctness.

%We have created a type system for determinism.  
We have created a type system that expresses determinism specifications
in a program.
The key ideas in the type system are type qualifiers for nondeterminism,
order-nondeterminism, and determinism; type well-formedness rules to
restrict the typings for collections; and enhancements to polymorphism that
improve precision when analyzing collection operations. While state-of-the-art
nondeterminism detection tools rely on observing runtime output, our approach
aims to verify determinism thereby providing stronger soundness guarantees.

We implemented our type system for Java.
Our type checker, \theDeterminismChecker, warns if a
program is nondeterministic or verifies that the program is deterministic.
In a case study of a 24,000 line software project, it found
previously-unknown nondeterminism errors in a program that had been heavily
vetted by its developers,
who were unusually concerned about nondeterminism errors.
In an experiment, it found all of the
non-concurrency-related nondeterminism that was found by
state-of-the-art dynamic approaches to detecting flaky tests.
